### 2020-03-30 01-introduction.tex
Update 01-introduction.tex

%Formally, Reaction Systems are defined as a rewriting formalism.
In this paper, we present an encoding from RSs, 
to the open multiparty process algebra c\CNA,\footnote{After `chained Core Network 
Algebra'.} a variant of the {\tt link}-calculus~\cite{BodeiBB12,BBB17} without name mobility.
This formalism allows several processes to synchronise and 
communicate altogether, at the same time, with a new communicating
mechanism based on links and link chains. 
Our initial motivation for introducing this mechanism was to encode
Mobile Ambients~\cite{CardelliG00}, getting a much stronger operational
correspondence than any available in the literature, such as the one in~\cite{B16}.
 This allowed us to easily encode calculi for biology equipped with membranes, as in~\cite{BodeiBBC14}.
Process calculi have been used successfully to model
biological processes, see~\cite{BBDFH18} for a recent survey.
We illustrate our embedding by means of some examples coming
both from the computer science and the biological field.
%and then we consider a more complex example, by modeling a RS 
%representing a regulatory network for \emph{lac} operon, presented in~\cite{CMMBM12}. 
We also show that our embedding preserves the main
features of RSs, and prove its correctness and completeness.
Then, we present a methodology for verifying formally
properties of reaction systems. The classical notion of bisimulation for 
process algebras allows to consider equivalent two processes
where one process can simulate all the actions executed 
by the other one and viceversa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to study for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
Besides bio-simulation, we define a completely new and simple assertion language,
which allows to specify properties of the reaction systems to be verified.
The classical logic of Hennessy-Milner~\cite{HM80} is a powerful
formalism for specifying properties of labeled transition systems, 
and we show that our assertions allow us to specify formulas 
that extend this logic to the case of complex labels.
Our formulas allow to abstract from some of the labels.

Our main contributions are as follows:
\begin{itemize}
\item the behaviour of the context, for each single 
entity,
%\footnote{Entities that belong to the `domain' of a RS.}
can be
specified in a recursive way as an ordinary process;
\item we can express the behaviour of entity mutation,
%change the behaviour of  each single entity $s$  in a
%way that a mutation $s'$ can be expressed; 
in such a way that the mutated entity
$s'$ can  take part to only a subset of rules requiring entity $s$;
\item with a little coding effort, two RSs can
communicate; i.e.  a subset of those entities that the context  can
provide, are then provided by a second RS;
\item as our translation results in  a  c\CNA \ system, from
each state only one transition can be generated, thus the c\CNA \
computation is fully deterministic.
\item we define an assertion language to specify local properties of 
reaction systems
\item we show that our assertions extend to our framework the Hennessy-Milner logic.
\end{itemize}

@@

As we have remarked, in our translation, Reaction Systems 
get the ability to interact between them in a synchronized manner. 
This interaction is not foreseen in
the basic RS framework, as it can only happen with the context.
By exploiting recursion, the kind of interactions which can be 
defined can be complex and expressive.
Example~\ref{ex:twoRS} and more in general the discussion in 
Section~\ref{sec:discussion} show that 
the interaction between RSs can help to model new scenarios.
\paragraph{Related work}
We are not aware of any inference rules for the RS; it seems not the case
that a deterministic behaviour, as the one of the RS, once the initial state is 
fixed, could be defined by inference rules that classically are applied 
to define non-deterministic transition systems.
The reversible computation paradigm for RS extends the RS framework  by
allowing  backward computations as well as forward computations; for this 
goal a set of inference rules for rewriting logic has been 
defined in~\cite{10.1007/978-3-319-73359-3_3}
to exactly keep trace of those elements that  dissolve in the next 
computation step.

>>>>

%Formally, Reaction Systems are defined as a rewriting formalism.
In this paper, we present an encoding from RSs, 
to the open multiparty process algebra c\CNA,\footnote{After `chained Core Network 
Algebra'.} a variant of the {\tt link}-calculus~\cite{BodeiBB12,BBB17} without name mobility.
This formalism allows several processes to synchronise and 
communicate altogether, at the same time, with a new interaction
mechanism based on links and link chains. 
The initial motivation for introducing this mechanism was to encode
Mobile Ambients~\cite{CardelliG00}, getting a much stronger operational
correspondence than any available in the literature, such as the one in~\cite{B16}.
Later it was shown that the {\tt link}-calculus allowed one to easily encode calculi for biology equipped with membranes, as in~\cite{BodeiBBC14}.

We illustrate our embedding by means of some examples coming
both from the computer science and the biological field.
%and then we consider a more complex example, by modeling a RS 
%representing a regulatory network for \emph{lac} operon, presented in~\cite{CMMBM12}. 
We also show that our embedding preserves the main
features of RSs, and prove its correctness and completeness.

Then, we present a methodology for verifying formally
properties of reaction systems. The classical notion of bisimulation for 
process algebras allows to consider equivalent two processes
where one process can simulate all the actions executed 
by the other one and viceversa. In this paper we define a new
notion of bisimilarity which takes into account the characteristics
of biological systems. We define a new {\em bio-simulation} having in mind
the possibility that two interacting systems may be compared w.r.t.
a subset of the possible biological actions. This is useful for concentrating 
on the sub-model that one may need to study for a specific study or application,
without getting lost in the complexity of the full biological system, or network.
The notion of bio-simulation relies on a new and simple assertion language,
which allows to focus on some properties of the reaction systems to be verified.
In fact, bio-similarity is parametric on a given assertion of interest.
The we prove that the well-known logical characterization of bisimilarity in terms of Hennessy-Milner Logic~\cite{HM80} (HML)
can be extended to the case of bio-similarity by tailoring HML formulas to
the same assertion of interest used in bio-similarity.  As HML is a powerful
formalism for specifying properties of labeled transition systems, 
its variant considered here looks a suitable option to specify formulas 
over complex labels by abstracting from unnecessary details.

Our main contributions are as follows:
\begin{itemize}
\item the behaviour of the context, for each single 
entity,
%\footnote{Entities that belong to the `domain' of a RS.}
can be
specified in a recursive way as an ordinary process;
\item along the same lines, non-deterministic contexts can be seamlessly expressed in our setting; 
\item when deterministic contexts are considered, the c\CNA \
computation is deterministic.
\item we define an assertion language to specify local properties of 
reaction systems and exploit it to define a novel notion of behavioural equivalence for reaction systems, called bio-similarity;
\item we show that our assertions extend can be used to extend the Hennessy-Milner logic to our framework, in such a way that logical equivalence of processes coincide with bio-similarity;
\item we sketch how one can express the behaviour of entity mutation,
%change the behaviour of  each single entity $s$  in a
%way that a mutation $s'$ can be expressed; 
in such a way that the mutated entity
$s'$ can  take part to only a subset of rules requiring entity $s$;
\item we show that with a little coding effort, two RSs can
communicate; i.e.  a subset of those entities that the context  can
provide, are then provided by a second RS;
\end{itemize}

@@

As we have remarked, in our translation, Reaction Systems 
get the ability to interact between them in a synchronized manner. 
This interaction is not foreseen in
the basic RS framework, as it can only happen with the context.
By exploiting recursion, the kind of interactions which can be 
defined can be complex and expressive.
Example~\ref{ex:twoRS} and more in general the discussion in 
Section~\ref{sec:discussion} show that 
the interaction between RSs can help to model new scenarios.

\paragraph{Related work}
Process calculi have been used successfully to model
biological processes, see~\cite{BBDFH18} for a recent survey.
We are not aware of any inference rules for the RS; it seems not the case
that a deterministic behaviour, as the one of the RS, once the initial state is 
fixed, could be defined by inference rules that classically are applied 
to define non-deterministic transition systems.
The reversible computation paradigm for RS extends the RS framework  by
allowing  backward computations as well as forward computations; for this 
goal a set of inference rules for rewriting logic has been 
defined in~\cite{10.1007/978-3-319-73359-3_3}
to exactly keep trace of those elements that  dissolve in the next 
computation step.
